---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r, echo = TRUE, message=FALSE}
library(dplyr)
library(lubridate)
#Loading in the proper data
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
download.file(url, 'activity.zip', method = 'curl')
unzip('activity.zip')
activity <- read.csv('activity.csv')
#removing NA's from data
activityNAomit <- activity[!is.na(activity$steps), ]
```
## What is mean total number of steps taken per day?
Histogram showing the distribution of steps per day 
```{r, echo=FALSE}
#Putting the total number of daily steps in a dataframe
totalDailySteps <- activityNAomit %>% group_by(date) %>% summarise(total = sum(steps))
#distribution of steps taken
hist(totalDailySteps$total, main = 'total daily steps distribution', xlab = 'number of steps')
```


5 number summary of steps per day
```{r, echo=FALSE}
#summary numbers for the total steps
summary(totalDailySteps$total)
```
## What is the average daily activity pattern?
```{r, echo=FALSE}
#Putting the total number of interval steps in a dataframe
meanTimeofDaySteps <- activityNAomit %>% group_by(interval) %>% summarise(mean = mean(steps))
#Putting the mean number of interval steps in a dataframe
plot(meanTimeofDaySteps$interval, meanTimeofDaySteps$mean, type = 'l', xlab = 'Interval (min)', ylab = 'Steps', main = 'Steps Vs Time', ylim = c(0, 250))
```


Finding the time of day when maximum steps usually occurs
```{r, echo=FALSE}
#Finding the 5 minute interval with the maximum average steps
timeMaxSteps <- meanTimeofDaySteps[meanTimeofDaySteps$mean == max(meanTimeofDaySteps$mean), 'interval']
```
**Time with the max steps is at `r timeMaxSteps` minutes **


## Imputing missing values
How many missing values exist for steps?:
```{r, echo=FALSE}
#finding how much data is not available
na_sum <- sum(is.na(activity$steps))
total_rows <- nrow(activity)
percent_missing <- round((na_sum/total_rows)*100, 2)
```
**There are `r na_sum` missing values for steps in our data out of `r total_rows` values, this means that `r percent_missing`% of values were missing**

When we impute the average activity level for all missing values instead of just ignoring them, we can see that this centralizes the distribution more

```{r, echo=FALSE}
#creating new activity df
imputed_mean <- activity
#replacing the steps with the average of steps
imputed_mean$steps[is.na(imputed_mean$steps)] <- mean(imputed_mean$steps, na.rm = TRUE)
totalDailyStepsimputed <- imputed_mean %>% group_by(date) %>% summarise(total = sum(steps))
hist(totalDailyStepsimputed$total, main = 'total daily steps distribution imputed', xlab = 'number of steps imputed')
```


The new 5 number summary after imputing the steps data
```{r, echo=FALSE}
summary(totalDailyStepsimputed$total)
```
## Are there differences in activity patterns between weekdays and weekends?
The weekends appear to have more activity in the middle of the day than the weekdays
```{r, echo=FALSE}
#dates casted as dates
imputed_mean$date <- ymd(imputed_mean$date)
imputed_mean$weekday <- wday(imputed_mean$date)
#finding mean steps at time of day for weekdays and weekends 
meanTimeofDayStepsWeekday <- imputed_mean[imputed_mean$weekday < 6, ] %>% group_by(interval) %>% summarise(mean = mean(steps))
meanTimeofDayStepsWeekend <- imputed_mean[imputed_mean$weekday >= 6, ] %>% group_by(interval) %>% summarise(mean = mean(steps))
#plotting mean steps at time of time of day for weekdays and weekends
par(mfrow = c(2,1))
plot(meanTimeofDayStepsWeekday$interval, meanTimeofDayStepsWeekday$mean, type = 'l', xlab = 'Interval (min)', ylab = 'Steps', main = 'Steps Vs Time Weekday', ylim = c(0, 250))
plot(meanTimeofDayStepsWeekend$interval, meanTimeofDayStepsWeekend$mean, type = 'l', xlab = 'Interval (min)', ylab = 'Steps', main = 'Steps Vs Time Weekend', ylim = c(0, 250))
```

